\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-11-19 honors_thesis/thesis/paper.tex
Some changes to my honors thesis based on a comments from Dr. Goodrich.  Not done yet.

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are common
place.  In order to efficiently utilize these resources for computationally
intensive problems, algorithms need to run in parallel.

@@

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function--additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

@@

While in some algorithms enumerating all possible next sampling positions would
be impossible or extremely difficult, in PSO all we need to know are the
current positions of the particle and its neighbors, from which we can easily
enumerate and evaluate all possible positions of that particle at the next
iteration.  The defining components of a particle's motion
(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The particle already
knows its personal best $\pbest$ at iteration $t-1$, but it could be updated to
$\ppos_t$ if its position at iteration $t$ has a better value than its current
personal best.  Likewise, the particle knows its neighborhood best $\nbest$ at
iteration $t-1$, though it could be updated by any of its neighbors' positions
at iteration $t$.

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has (a particle either updates its personal best or it
does not, and one of its $n$ neighbors must update its neighborhood best, or
there is no neighborhood best update--$n+1$ possibilities).  If all of those
speculative positions are evaluated in parallel along with the original,
non-speculative position of the particle at time $t-1$, then one of the
speculative positions has to be the correct one, and two iterations have been
evaluated at once.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
$\sset_{t+1}$.  $\n_t$ is a neighbor of particle $\p_t$.  A set of particles is
$\pset$, whereas a single particle is simply $\p$.

@@

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
personal bests, but $\p_t$'s value was better.  For example, suppose that
$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
best with its position at time $t$.  Then both $\noeval{\p}_t$ and
$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
$\nonbest{\s}_{t+1}$ would think that its current position was its personal
best, as the value it found--2--was better than its previous personal best
value of 3.  It needs to receive the personal best value from its parent to
know that its personal best position $\pbest$ is actually the position of
$\p_t$, not $\s_{t+1}$.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for the particles it
manages.  Messages such as values and positions for the neighborhood best are
sent between processors.  There may be (in fact there probably must be) some
machine that collects information from all of the particles and outputs the
result of the algorithm, though that machine's importance is much less than in
centralized algorithms.

To perform speculative evaluation in a distributed PSO algorithm, there must be
some way to have processors evaluate the speculative children of particles,
without giving the speculative particles the same treatment as actual
particles.  Perhaps that can be done by assigning each particle a set of
machines instead of a single machine, and the particle directs its extra
machines to evaluate its speculative children.  In the MapReduce framework,
each particle and speculative child has its own Map task, during which the
particle is evaluated, with the speculative children sending their output back
to the original particle's Reduce task and any neighbors it has.  The Reduce
task for each particle gets all of the messages from its neighbors, its
speculative children, and the speculative neighbors of its speculative
children, and produces the next particle and children for another Map task.

No matter which framework is used the message passing should be similar, and
that is what we describe here.

@@

This second method only requires one round of passing information because
information about iteration $t$ and $t+1$ are passed at the same time.  Each
processor reconstructs from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children--neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same information for
the neighboring particle that we needed to produce the original particle,
$\nonbest{\p}_{t+1}$--we need the original neighbor particle, its speculative
children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  With $\nonbest{\nset}_{t+1}$ and
$\nonbest{\p}_{t+1}$, we can produce $\p_{t+1}$.

@@

It has been shown that in dynamic topologies using the $\nbest$ only from
current neighbors leads to poor performance--$\nbest$ should instead be defined
as the best position of any of the neighbors the particle has ever
had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
is an open one.  Giving more information to a particle's neighbors shares
information more quickly through the swarm.  We choose to share the personal
best value as a good middle ground.  Also, sharing the neighborhood best would
require more message passing in distributed algorithms, as particles would need
more information in order to generate their speculative children from their
neighbors.

Performing speculative evaluation in PSO with a dynamic topology also raises a
few issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors--it needs to send messages
to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.  For
every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random.

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation--there are potentially
more than $2(n+1)$ possible next positions, increasing the amount of work that
must be done to perform the second iteration at the same time as the first.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  Then only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
that SEPSO clearly beat PSO with a Random topology.  However, the performance
between SEPSO and PSO Complete is not as encouraging--the results are
indistinguishable.  But SEPSO Pick Best handily outperforms all other
algorithms in this problem.

@@

To further belabor the point that performing speculative evaluation is a much
better idea with some functions than simply adding particles, in \fig{sphere2}
we show a graph comparing the performance of SEPSO Pick Best with 13 particles
and with 50 particles.  The difference between the two in terms of function
evaluations is staggering.

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely add more particles to
the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using this
algorithm, the original PSO algorithm is exactly reproduced two iterations at a
time.  This technique requires more function evaluations per iteration than
regular PSO, but for a number of functions still performs better.

@@

Still another possibility is to view the set of possible evaluations as an
infinite tree.  Each level in the tree is an iteration, and the branches on
each level represent the possible next positions of the particle.  One need not
simply evaluate all of one level for each particle--perhaps going two or three
iterations ahead on the most likely path would perform better.

>>>>

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are
commonplace.  In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

@@

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

@@

While in some algorithms enumerating all possible next sampling positions would
be impossible or extremely difficult, in PSO all we need to know are the
current positions of the particle and its neighbors, from which we can easily
enumerate and evaluate all possible positions of that particle at the next
iteration.  The defining components of a particle's motion
(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The only remaining
components are $\pbest$ and $\nbest$, each of which has only a few
possibilities.  The particle already knows its personal best $\pbest$ at
iteration $t-1$, but it could be updated to $\ppos_t$ if its position at
iteration $t$ has a better value than its current personal best.  Likewise, the
particle knows its neighborhood best $\nbest$ at iteration $t-1$, though it
could be updated by any of its neighbors' positions at iteration $t$.

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has, because there are two possibilities for updating a
particle's personal best (updated and not updated) and $n+1$ possibilities for
updating its neighborhood best (one of $n$ neighbors, or no update).  If all of
those speculative positions are evaluated in parallel along with the original,
non-speculative position of the particle at time $t-1$, then one of the
speculative positions has to be the correct one, and two iterations have been
evaluated at once.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children.
$\n_t$ is a neighbor of particle $\p_t$.  A set of particles is $\pset$,
$\sset$, or $\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.

@@

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
personal bests, but $\p_t$'s value was better.  For example, suppose that
$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
best with its position at time $t$.  Then both $\noeval{\p}_t$ and
$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
$\nonbest{\s}_{t+1}$ would think that its current position was its personal
best, as the value it found, 2, was better than its previous personal best
value of 3.  It needs to receive the personal best value from its parent to
know that its personal best position $\pbest$ is actually the position of
$\p_t$, not $\s_{t+1}$.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for the particles it
manages.  Messages such as values and positions for the neighborhood best are
sent between processors.  There may be some machine that collects information
from all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

To perform speculative evaluation in a distributed PSO algorithm, there must be
some way to have processors evaluate the speculative children of particles,
without giving the speculative particles the same treatment as actual
particles, as the speculative children only live for one iteration.  Perhaps
that can be done by assigning each particle a set of machines instead of a
single machine, and the particle directs its extra machines to evaluate its
speculative children.  No matter the distributed framework used the message
passing should be similar, and that is what we describe here.

@@

This second method only requires one round of passing information because
information about iteration $t$ and $t+1$ are passed at the same time.  Each
processor reconstructs from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children---neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same information for
the neighboring particle that we needed to produce the original particle,
$\nonbest{\p}_{t+1}$; we need the original neighbor particle, its speculative
children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  With $\nonbest{\nset}_{t+1}$ and
$\nonbest{\p}_{t+1}$, we can produce $\p_{t+1}$.

@@

It has been shown that in dynamic topologies using the $\nbest$ only from
current neighbors leads to poor performance.  $\nbest$ should instead be
defined as the best position of any of the neighbors the particle has ever
had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
is an open one.  Giving more information to a particle's neighbors shares
information more quickly through the swarm.  We choose to share the personal
best value as a good middle ground.  Also, sharing the neighborhood best would
require more message passing in distributed algorithms, as particles would need
more information in order to generate their speculative children from their
neighbors.

Performing speculative evaluation in PSO with a dynamic topology also raises a
few issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random.

@@

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2(n+1)$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches that branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The branch that PSO would have taken is ignored, but $\p_t$ must still be
created so it can give the right personal best and neighborhood best values and
positions to $\p_{t+1}$.  The only change needed in \alg{centralized} is in
step 7, where the $\noeval{\s}_{t+1}$ with the highest value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
that SEPSO clearly beat PSO with a Random topology.  However, the performance
between SEPSO and PSO Complete is not as encouraging---the results are
indistinguishable.  But SEPSO Pick Best handily outperforms all other
algorithms in this problem.

@@

To further illustrate the point that performing speculative evaluation is a
much better idea with some functions than simply adding particles, in
\fig{sphere2} we show a graph comparing the performance of SEPSO Pick Best with
13 particles and with 50 particles.  The difference between the two in terms of
function evaluations is staggering.  However, this result is well known; with
Sphere and similar functions, the best thing to do is have a small swarm run
longer.  We show the results here because our method allows the algorithm to
use spare processors in parallel to perform more iterations at the same time.
Clearly with these functions the best thing to do is not to add particles to
the swarm, but to perform speculative evaluation.

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using this
algorithm, the original PSO algorithm is exactly reproduced two iterations at a
time.  This technique requires more function evaluations per iteration than
regular PSO, but for a number of functions still performs better.

@@

Still another possibility is to view the set of possible evaluations as an
infinite tree.  Each level in the tree is an iteration, and the branches on
each level represent the possible next positions of the particle.  One need not
simply evaluate all of one level for each particle---perhaps going two or three
iterations ahead on the most likely path would perform better.
